---
title: "Monte Carlo Integration, Variance and Efficiency"
subtitle: "Chapter 6 SCR2e, Part II, 6.3-6.8"
author: "AG Schissler"
date: "4 Mar 2021 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
set.seed( 03022021 )
library(tidyverse)
library(printr)
## xaringan::inf_mr('11_ch6_mcIntegrationVarReduce2.Rmd')
## str(knitr::opts_chunk$get())
```

<!-- open up etext as you go. don't typeset the chapter -->

# i. Admin & Startup

## Schedule review

- Grades up to date. Sorry for the delay. Any questions that are appropriate for the whole group?
- Let's review the schedule and place today's learning in context.  
- Please complete Ch. 5 Problem Set by Friday. 

##  Questions regarding Ch. 5 HW?

- 5.3-5.6 Two-component mixture viz (vary the mixing parameter)
- 5.7 (parallel coordinates)
- 5.11 (segment-style stars plot)
- 5.12, 5.13 Understanding PCA
- 5.16 Applying PCA

## Today's plan

i. Finish Discussion of Section 6.2, with Examples 6.4 and 6.5.
ii. Preview Ch. 6 Problem Set questions to prime learning.
iii. Discuss Ch. 6, Part II: Variance reduction techniques
iv. Summarize, conclude, revisit Ch. 6 Problem prompt discussion / hints time remaining

*Please contribute questions and comments*

# Finish Discussion of Section 6.2, with Examples 6.4 and 6.5.

## Example 6.4

```{r ex64}
### Example 6.4 (Example 6.3, cont.)
x <- seq(.1, 2.5, length = 10)
m <- 10000
z <- rnorm(m)
dim(x) <- length(x)
p <- apply(x, MARGIN = 1,
           FUN = function(x, z) {mean(z < x)}, z = z)
```

---

```{r ex64b}
Phi <- pnorm(x)
print(round(rbind(x, p, Phi), 3))
```


## Example 6.5

```{r ex65}
### Example 6.5 (Error bounds for MC integration)
x <- 2
m <- 10000
z <- rnorm(m)
g <- (z < x)  #the indicator function
v <- mean((g - mean(g))^2) / m
cdf <- mean(g)
c(cdf, v)
c(cdf - 1.96 * sqrt(v), cdf + 1.96 * sqrt(v))
pnorm(2)
```

## 6.2.2 Variance and Efficiency

- discuss p.153

# II. Preview Ch.6 Problem Set (quickly)

- 6.1 (MC integration; see example 6.1,6.2)
- 6.2 (MC `pnorm`; see example 6.3)
- 6.3 (comparing MC estimators)
- 6.6, 6.7 (antithetic variates)
- 6.12 (importance sampling theory)
- 6.13 (importance sampling)
- 6.15 (stratified importance sampling; see examples 6.11 and 6.14)

# iii. Discuss Ch. 6, Part II: Variance reduction techniques

# 6.4 Antithetic variables

## Proposition 6.1

- Sketch the proof.
- Any questions?

## Corollary 6.1

Core insight is:

If $U \sim U(0,1)$, then $1-U \sim U(0,1)$ but clearly

$$
Cov(U, 1-U) = Cov(U, -U) + Cov(U, 1) = - Cov(U,U) + 0 = - Var(U) < 0
$$

And applying $g( F_X^{-1}( \cdot ) )$ doesn't mess this up as the probability of integral of both $g( F_X^{-1}( U ) )$ and $g( F_X^{-1}( 1-U ) )$ get us to $g(x)$

## Example 6.6 (Antithetic variables)

```{r 66a}
MC.Phi <- function(x, R = 10000, antithetic = TRUE) {
    u <- runif(R/2)
    if (!antithetic) v <- runif(R/2) else
                                         v <- 1 - u
    u <- c(u, v)
    cdf <- numeric(length(x))
    for (i in 1:length(x)) {
        g <- x[i] * exp(-(u * x[i])^2 / 2)
        cdf[i] <- mean(g) / sqrt(2 * pi) + 0.5
    }
    cdf
}
```

## Example 6.6 (Antithetic variables)
    
```{r 66b}
x <- seq(.1, 2.5, length=5)
Phi <- pnorm(x)
set.seed(123)
MC1 <- MC.Phi(x, anti = FALSE)
set.seed(123)
MC2 <- MC.Phi(x)
print(round(rbind(x, MC1, MC2, Phi), 5))
```

## Example 6.6 (Antithetic variables)
    
```{r 66c}
m <- 1000
MC1 <- MC2 <- numeric(m)
x <- 1.95
for (i in 1:m) {
    MC1[i] <- MC.Phi(x, R = 1000, anti = FALSE)
    MC2[i] <- MC.Phi(x, R = 1000)
}
print(sd(MC1))
print(sd(MC2))
print((var(MC1) - var(MC2))/var(MC1))
```

# 6.5 Contral Variates

## 6.5 Contral Variates background

- Discuss p.158-159.

## Example 6.8 (MC integration using control variates)

```{r 68a}
f <- function(u)
    exp(-.5)/(1+u^2)

g <- function(u)
    exp(-u)/(1+u^2)

set.seed(510) #needed later
u <- runif(10000)
B <- f(u)
A <- g(u)
cor(A, B)
```

## Example 6.8 (MC integration using control variates)

```{r 68b}
a <- -cov(A,B) / var(B)    #est of c*
a
```

## Example 6.8 (MC integration using control variates)

```{r 68c}
m <- 100000
u <- runif(m)
T1 <- g(u)
T2 <- T1 + a * (f(u) - exp(-.5)*pi/4)
c(mean(T1), mean(T2))
c(var(T1), var(T2))
(var(T1) - var(T2)) / var(T1)
```

## 6.5.1 Antithetic Variate as Control Variate

Discussion

# 6.6 Importance sampling

- Introduction on p. 167.

## Example 6.11, 6.12

From Rizzo's repo.

# 6.7 Stratified Sampling

- Introduction on p. 172.

## Proposition 6.2

Stratification helps when means are the strata vary.

## Example 6.13

From Rizzo's repo.

# 6.8 Stratified Importance Sampling

## Proposition 6.3




# Closing / Ch. 6 Problem set

- Revisit HW.
- Questions?
- Students please summarize/ask questions about what we discussed today.
