---
title: "PERMUTATION TESTS"
subtitle: "Chapter 10 SCR2e, Part 2"
author: "AG Schissler"
date: "1 Apr 2021 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
set.seed( 04012021 )
library(tidyverse)
library(printr)
## xaringan::inf_mr('17_ch10_PermutationTests1.Rmd')
## str(knitr::opts_chunk$get())
```

<!-- open up etext as you go. don't typeset the chapter -->

# i. Week 10 Focus

- Students Practice on Ch. 8, 9 content.
- Students receive and act on Ch. 7 feedback.
- Read Permutation Tests Ch. 10.
- Discuss Permutation Tests Ch. 10.

## Any questions on Ch.8, 9 HW?

- 8.7 - 8.11
- 9.4 - 9.6

## Today's plan

i. Open with questions on Ch. 10 Problem Set to prime learning.
ii. Discuss Ch. 10 Problems
iii. Summarize, conclude, revisit Ch. 10 Problem prompt discussion / hints time remaining

*Please contribute questions and comments*

# Ch.10 Permutation Tests

- 10.1 Introduction
- 10.2 Tests of Equal Distributions
- 10.3 Multivariate Tests for Equal Distributions
- 10.4 Application: Distance Correlation

# I. Ch.10 HW

- 10.1 - 10.3, 10.6

# 10.3 Multivariate Tests for Equal Distributions

- 10.3.1 Nearest Neighbor Tests
- 10.3.2 Energy Test for Equal Distributions

## 10.3.1 Nearest Neighbor Tests

- Notation/Discussion
- Found it is interesting that 

> "In general, if the sampled distributions are equal, the pooled sample has on average less nearest neighbor coincidences than under the alternative hypothesis".

Counterintuitive at first... 

## Remark 10.1 

- This remark is interesting on the motivation behind why choose permutation test opposed to bootstrap.
- Ties are always a consideration in probability calculation
- Ties occur for discrete rvs and resampled values (same values chosen)
- Assuming continous motivates sampling *without replacing* in the permutation space.

##  Using `yaImpute`.

```{r yaImpute}
library(yaImpute)
help( package = 'yaImpute' )
```

##  Example 10.4

Finding nearest neighbors

##  Example 10.5

Nearest neighbor statistic

##  Example 10.6

Nearest neighbor test

## 10.3.2 Energy Test for Equal Distributions

## Example 10.7

Two-sample energy statistic

## Note `energy` package

- Use `energy` package for optimized code (instead of Rizzo's implementation).
- In general, finding a good package for your research is a good idea (or making one if there is a gap!)

## Example 10.8 

Two-sample energy test

## Example 10.9 

k-sample energy distances

## Example 10.10

Distance Components (disco)

## Example 10.11

Power Comparison

# 10.4 Application: Distance Correlation

- Interesting dependence measure, very general.
- Definition 10.1 empirical distance covariance 
- Definition 10.2 empirical distance correlation

## Example 10.12 

Distance covariance statistic

## Example 10.13 

Distance correlation statistic

## Example 10.14 

Distance covariance test

# Closing / Ch. 10 Problem set

- Revisit HW.
- Questions?
- Students please summarize/ask questions about what we discussed today.
