---
title: "Introduction to statistical computing in `R` (Ch. 1)"
author: "AG Schissler"
date: "28 Jan 2021 (*updated: `r Sys.Date()`)*"
output:
  html_document
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(printr)
## library(rstanarm)
## options(mc.cores = parallel::detectCores() - 1)
set.seed(02162021)
## xaringan::inf_mr('2_tools_workflows.rmd')
## system( 'cp ~/bib/CP3.bib ../.' )
knitr::opts_chunk$set(cache=FALSE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%')
## str(knitr::opts_chunk$get())
```

<!-- open up etext as you go. don't typeset the chapter -->

# 1. Introductory remarks on statistical computing

## 1.1 Statistical computing

- Assess bias and efficiency of estimators at finite sample sizes
- Alternative to asymptotic arguments when intractable or undesired (non-asymptotic viewpoint).
- Big Data requires additional computing and visualization techniques

- `R` gives us unique access to great simulation tools (unique compared to other languages). Why simulate? Welcome to the 21st century! Two reasons:
  1) Often, simulations can be easier than hand calculations 
  2) Often, simulations can be made more realistic than hand calculations.

- automation and reproducibility

# 2. Introduction to `R` 

## The `R` computing environment

- Everyone got `R` up and running?
- Any issues you'd like to discuss now?

## Getting started with `R` and `Rstudio`

- Let me demonstrate rendering these slides.
- Please download this rmd from Canvas Files and render
- Any issues?

## Basic syntax

- Let's look at Rizzo Section 1.4.
- Be sure to ask if you have any questions as we go.

## Using the `R` online help system

```{r coefHelp}
?coef
```

## CRAN / CRAN task view

CRAN

```{r cranInstall, eval=F}
install.packages("printr")
```

-CRAN Task  view

https://cloud.r-project.org/web/views/

## Git and github-hosted R packages

Let's retrieve Rizzo's scripts:

```{bash cloneGit, eval=F, include=T}
git clone https://github.com/mariarizzo/SCR.git
```

https://github.com/yihui/printr

https://github.com/SchisslerGroup/bigsimr/tree/master

writing your own package:

https://r-pkgs.org/

## Vignettes

Let's walk through the `printr` vignette:

https://cran.r-project.org/web/packages/printr/vignettes/printr.html

# 3. Basic use of `R`

## Distributions 

```{r distr}
help.search("distribution", package="stats")
```

## Statistical tests

```{r tests}
help.search(".test", package="stats")
```

## Using scripts

Lets' open Rizzo's `SCR/SCR-2e/R/SCR2e-Intro.R` in Rstudio.

## Functions

- Let's run the Section 1.7 code interactively.

## Arrays, data frames, lists

- Let's run the Section 1.8 code interactively.

## Formulas for shorthand regression specification

```{r formulaExamples}
example(formula)
```

# 4. Data viz, project management, and workflow/report generation

## Graphics

https://www.r-graph-gallery.com/

## `ggplot`

https://www.r-graph-gallery.com/ggplot2-package.html

https://ggplot2.tidyverse.org/

graphic example for our research:  
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4908332/

## Workspace and files

- Quick tour of Rstudio
- https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects

## Using `rmarkdown` and `knitr`

https://github.com/grizant/STAT745/blob/master/meeting_agendas/2_tools_workflows.Rmd

```{r render, eval=F}
## if file is downloaded to current directory
rmarkdown::render( '2_tools_workflows.Rmd' )
```
