---
title: "Probability and Statistics Review (Ch. 2)"
subtitle: "Part I"
author: "AG Schissler"
date: "2 Feb 2021 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(printr)
## xaringan::inf_mr('3_prob_stat_review1_slides.rmd')
## str(knitr::opts_chunk$get())
```

<!-- open up etext as you go. don't typeset the chapter -->

# i. Admin

## Problem Set 1 details and rubric

https://unr.canvaslms.com/courses/63835/assignments/766606

## Any questions about course administration/last lecture?

Please ask now.

## Today's plan

- Let's go through ch. 2 and I'll make comments.  
- Please contribute questions, comments, etc.

# I. Prob review

## Course Prerequisites and expected student preparation

- STAT 667 or STAT 754 or STAT 726.  
- Recommended Preparation: STAT 645; MATH 330.  
- How did this chapter read for you?  


## 2.1 Random variables and probability

- Pretty standard definitions.  
- joint probability $P( A \cap B )$ notated as $P( AB )$.  
- If $X$ and $Y$ are independent, then $Cov(X,Y)=\rho(X,Y)=0$. But the converse is not true in general.   
- An important special case is that for **normal** rv's if $Cov(X,Y)=\rho(X,Y)=0$ then $X$ and $Y$ are independent. (p. 41).

# II. Discrete

## 2.2 Some discrete distributions

- notice the support of the alternative formulation of geometric.  
- briefly discuss the techniques used in Examples 2.1 and 2.2.  

## Be sure to check what parameterization `R` is using

```{r geom}
?dgeom
```

# III. Continuous

## 2.3 Some continuous distributions

```{r norm}
?dnorm
```

## continuous examples

- Discuss Example 2.3  
- Discuss Example 2.4

# IV. Multivariate distributions 

## 2.4 Multivariate normal

- powerful, flexible framework for correlated data modeling

## efficient, multicore multivariate normal simulation

```{r mvnfast}
library(mvnfast)
help(package=mvnfast)
```

## `mvnfast` random mvn vectors

```{r rmvn}
example(rmvn)
```

# V. Limit Theorems

## Limit theorems

## Relationships among prob distributions

Let's look at this and discuss:

https://en.wikipedia.org/wiki/Relationships_among_probability_distributions

# Closing

- Questions?
- 3-2-1.
