---
title: "Visualization of Multivariate Data"
subtitle: "Chapter 5 SCR2e"
author: "AG Schissler"
date: "23 Feb 2021 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
set.seed( 02232021 )
library(tidyverse)
library(printr)
## xaringan::inf_mr('9_ch5_multivariateViz.Rmd')
## str(knitr::opts_chunk$get())
```

<!-- open up etext as you go. don't typeset the chapter -->

# i. Admin & Startup

## Schedule review

- Let's review the schedule and place today's learning in context.  
- Please complete the **midterm exam** and Ch. 4 Problem Set by Friday. 
- Questions regarding Ch. 4 HW?
- Clarification questions for the midterm exam?

## Today's plan

I. Preview Ch. 5 Problem Set questions to prime learning.  
II. Discuss Ch. 5  
III. Revisit Ch. 5 Problem prompt discussion / hints    

*Please contribute questions and comments*

# I. Preview Ch.5 Problem Set

- 5.3-5.6 Two-component mixture viz (vary the mixing parameter)
- 5.7 (parallel coordinates)
- 5.11 (segment-style stars plot)
- 5.12, 5.13 Understanding PCA
- 5.16 Applying PCA

# II. Chapter 5 discussion

## 5.1 Introduction

- Importance of viz for exploratory data analysis (EDA).
- The term “exploratory” is in contrast to “confirmatory,” which could describe hypothesis testing. 
- Tukeybelieved that it was important to do the exploratory work before hypothesis testing, to learn what are the appropriate questions to ask, and the most appropriate methods to answer them.
- With multivariate data, we may also be interested in dimension reduction or finding structure or groups in the data. Here we restrict attention to methods for visualizing multivariate data.
- Discuss Table 5.1

# Sections 5.2 - 5.5

## 5.2 Panel displays

- Consider using `GGally::pairs` for scatterplot matrix.
- Discuss Example 5.1.

## 5.3 Correlation plots

- dimension reduction
- useful for predictive modeling
- fitting considerations for highly correlated features
- Discuss Example 5.2 (Decathlon data)

## 5.4 Surface plots and 3D Scatter plots

- Discuss Example 5.3
- `expand.grid` is great for simulation configurations
- Discuss Figure 5.7 scatter plot briefly 

## 5.5 Contour Plots

- briefly discuss in Rizzo.

# 5.6 Other 2D Representations of Data

## Example 5.11 (Parallel Coordinate)

```{r ex511}
library(MASS)
library(lattice)
##trellis.device(color = FALSE) #black and white display
x <- crabs[seq(5, 200, 5), ]  #get every fifth obs.
parallelplot(~x[4:8] | sp*sex, x)
```

---

```{r ex511b}
##trellis.device(color = FALSE)    #black and white display
x <- crabs[seq(5, 200, 5), ]     #get every fifth obs.
a <- x$CW * x$CL                 #area of carapace
x[4:8] <- x[4:8] / sqrt(a)       #adjust for size
parallelplot(~x[4:8] | sp*sex, x)
```

## 5.6.3 Segments, stars, and other representations

From Gardeux, Vincent, et al. "‘N-of-1-pathways’ unveils personal deregulated mechanisms from a single pair of RNA-Seq samples: towards precision medicine." Journal of the American Medical Informatics Association 21.6 (2014): 1015-1025.
https://academic.oup.com/jamia/article/21/6/1015/2909265?login=true

## Example 5.12 (segment plot)

```{r ex512}
x <- MASS::crabs[seq(5, 200, 5), ]         #get every fifth obs.
x <- subset(x, sex == "M")           #keep just the males
a <- x$CW * x$CL                     #area of carapace
x[4:8] <- x[4:8] / sqrt(a)           #adjust for size

palette(rainbow(6))                 #or use color
stars(x[4:8], draw.segments = TRUE,
      labels =levels(x$sp), nrow = 4,
      ylim = c(-2,10), key.loc = c(3,-1))

palette("default")
```

## Single-cell transcriptome dysregulation viz

- Polar-coordinate bar chart from our research:  

Schissler AG, Li Q, Chen JL, Kenost C, Achour I, Billheimer DD, Li H, Piegorsch WW, Lussier YA. *Analysis of aggregated cell-cell statistical distances within pathways unveils therapeutic-resistance mechanisms in circulating tumor cells. Bioinformatics.* 2016 Jun 15;32(12):i80-i89. doi: 10.1093/bioinformatics/btw248. PMID: 27307648; PMCID: PMC4908332.
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4908332/

- `nof1` pathways R package vignette: http://www.lussierlab.net//publications/nof1pathways-vignette_v2.html

Ask me if you want the code, it is under patent.

# 5.7 Principal Component Analysis (PCA)

## 2D PCA illustration

https://en.wikipedia.org/wiki/Principal_component_analysis#/media/File:GaussianScatterPCA.svg

## Example 5.13 (PCA)

```{r ex513a}
library(bootstrap)
str(scor)
```

---

```{r ex513b, message=FALSE, warning=FALSE}
library(GGally)
plot513b <- ggpairs(scor)
plot513b
```

---

```{r ex513c}
n <- nrow(scor)
x <- scale(scor)  #center and scale
s <- cov(x)
e <- eigen(s)
lam <- e$values   #vector of eigenvalues
P <- e$vectors    #matrix of eigenvectors
## plot(lam, type = "b", xlab = "eigenvalues", main = "")
barplot(lam, xlab = "eigenvalues")
```

---

```{r ex513d, message=FALSE, warning=FALSE}
tab <- rbind(lam / sum(lam), cumsum(lam) / sum(lam))
tab
```

---

```{r ex513e, message=FALSE, warning=FALSE}
z <- x %*% P
dim(z)
head(z)
```

---

```{r ex513f}
pc <- prcomp(scor, center = TRUE, scale = TRUE)
summary(pc)
```

---

```{r ex513g}
df <- scor[1:5, ]
predict(pc, newdata = df)  #same as z above
```

---

```{r ex513h}
head(x %*% pc$rotation, 5)
```

---

```{r ex513i}
head(pc$rotation)
```

---

```{r ex513j, warning=F}
head(P)
```

## Example 5.14 (PC Biplot)

```{r ex514}
## plot scor data in the (PC1, PC2) coordinate system
biplot(pc, pc.biplot = TRUE)
round(cor(x, z), 3)
```

## Example 5.15 ( Decathlon PCA)

### Example 5.15 (PCA: Decathlon data)

```{r ex515}
library(FactoMineR)
data(decathlon)
pc <- princomp(decathlon[,1:10], cor = TRUE, scores = TRUE)
plot(pc)     # screeplot
```

---
    
```{r ex515b}
biplot(pc)
```

---
    
```{r ex515c}
summary(pc)    
```

# 5.8 Other approaches to Data Visualization

- interesting content such as *grand tour* viz via `tourr` package.
- `FactoMineR` for multivariate viz
- https://datavis.ca/ categorical data
- `forcats` in tidyverse


# 5.9 Additional resources

Good list of resources. Check them out for inspiration during when developing your next important plot.

# Closing

- Students please summarize/ask questions about what we discussed today.
- Other questions?
