---
title: "STAT 745 Statistical Computing"
author: "AG Schissler"
date: "Spring 2021"
output: 
  html_document:
    toc: yes
---

# Basic Information

* Room number: DMS 106
* Meeting times: TuTh 4:30pm to 5:45pm.
* Contact information: [mailto:aschissler@unr.edu](aschissler@unr.edu), 775-784-4661 (office)
* Office room number: DMS 224
* Office hours: MoTu 4-5pm and by appointment

# Description


# Catalog Description

Computing using statistical software. Resampling and simulation: bootstrap, Monte Carlo, and Markov chain methods. Numerical inference for non-standard models. Bayesian inference. Advanced graphics

# Course Pre-requisites

STAT 667 or STAT 754 or STAT 726; Recommended Preparation: STAT 645; MATH 330.

# Catalog Student Learning Outcomes

Upon completion of this course, students will be able to:  

1. organize and present information in graphical and numerical forms to support concise, comprehensible, and scientifically defensible written conclusions/knowledge to research problems.  
2. create computational solutions to answer statistical questions for traditional and non-standard models using a high-level programming language.  
3. organize computational research findings in a professional report suitable for presentation to a variety of audiences  

# Textbooks

## Required 

Rizzo

## Supplemental texts

# Assignments

## Exercises

Exercises will be assigned approximately weekly. You are encouraged to discuss assignments between each other and with instructor. However, the assignment must be completed and submitted individually. 

## Midterms
There will be three midterms, the first on Wednesday, February 5, the second on Wednesday, March 11, and the third on Wednesday, April 8.

## Final exam
Due to the computational complexity involved in latter topics, the final exam will be take home.

## Exam policy

You will be allowed at one 8.5x11in page of handwritten (on both sides) notes for the first midterm. And one additional page of notes for each following midterm (three total pages). If you believe that your grade for exam or assignment is incorrect, contact instructor at the office hours with a rational justification. All such requests must be submitted to instructor within one week after a grade is announced; late requests will not be granted. Please understand that everyone can make a mistake, and that mistakes can go both ways: higher or lower than original grade.

## Makeup, Late Policy
Late assignments, exams, and projects will be graded with a 5 point (out of 20) penalty per day. Exceptions will be made when a student misses work due to a documented (doctor's note) illness or an extraordinary situation (up to the discretion of the instructor). There will be no early or make-up exams. However, if you need to miss an exam due to participation in a religious holiday or an official university activities (including athletics and other sanctioned activities), you must make arrangements with the instructor at least two weeks prior to the exam in question. **I'll drop your lowest grade in the ``Assignments'' category to give a more realistic description of your performance, if you aren't able to finish all assignment on time.**

# Grading

We'll use a point system to evaluate student learning. There are four categories of assignments. Table XXX shows the total points possible within each category before dropping any low scores.

```{r grading}

```

The instructor reserves the right to deviate from the above thresholds, including assigning $+$ or $-$.

# Diversity Statement

The University of Nevada, Reno is committed to providing a safe learning and work environment for all. If you believe you have experienced discrimination, sexual harassment, sexual assault, domestic/dating violence, or stalking, whether on or off campus, or need information related to immigration concerns, please contact the University’s Equal Opportunity & Title IX Office at (775) 784-1547. Resources and interim measures are available to assist you. For more information, please visit [http:www.unr.edu/equal-opportunity-title-ix](http:www.unr.edu/equal-opportunity-title-ix).

# Disability Statement
Any student with a disability needing academic adjustments or accommodations is requested to speak with the [http:www.unr.edu/drc](Disability Resource Center) as soon as possible to arrange for appropriate accommodations.

# Academic Conduct
No laptops, cell phones, mp3 players, or other electronics are to be used for personal reasons in class. If you are being disruptive during class you will be asked to leave. Disruptions in this context include inadequate participation. You must come to class on time and stay until the end of lecture. Tardy students will not be admitted to class. Please visit [http:www.unr.edu/student-conduct](http:www.unr.edu/student-conduct) for our official student code of conduct.

# Academic Success Services
Your student fees cover usage of the University Math Center, University Tutoring Center, and University Writing Center. These centers support your classroom learning; it is your responsibility to take advantage of their services. Keep in mind that seeking help outside of class is the sign of a responsible and successful student.

# University Recording Policy
Surreptitious or covert videotaping of class or unauthorized audio recording of class is prohibited by law and by Board of Regents policy. This class may be videotaped or audio recorded only with the written permission of the instructor. In order to accommodate students with disabilities, some students may have been given permission to record class lectures and discussions. Therefore, students should understand that their comments during class may be recorded.

# Academic Dishonesty
Cheating, plagiarism, or otherwise obtaining grades under false pretenses constitutes academic dishonesty according to the code of this university. Academic dishonesty will not be tolerated and penalties can include canceling a student’s enrollment without a grade or giving an F for the assignment or for the entire course. See the University Academic Standards policy: [UAM 6,502](https://www.unr.edu/administrative-manual/6000-6999-curricula-teaching-research/instruction-research-procedures/6502-academic-standards).

# Tentative course schedule

```{r schedule}

```
